Towards a Wide-Coverage Grammar: \\ Graphical Abstract Categorial Grammars

### 2013-07-08 lenls/abstract.tex
Framework for LENLS abstract
First draft of LENLS abstract

\documentclass{report}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{gb4e}
\usepackage{bussproofs}
\usepackage{hyperref}
\usepackage[labelformat=simple]{subcaption}
\renewcommand\thesubfigure{(\alph{subfigure})}
\usepackage{cancel}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage{pdfpages}
\usepackage{enumitem}

@@

\pagenumbering{gobble}
\includepdf[pages={1}]{front-page.pdf}
\newpage
~                               %
\newpage

\input{abstract}
\input{acknowledgments}

\tableofcontents

\chapter{Preliminaries}
\pagenumbering{arabic}
\setcounter{page}{1}

\input{motivation}
\input{outline}
\input{acg}
\input{ig}

\input{implementation}

\input{constraints}

\input{gacg}

\input{conclusion}

\bibliographystyle{plain}
\bibliography{biblio}

>>>>

\documentclass[twocolumn]{article}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{gb4e}
\usepackage{bussproofs}
\usepackage{hyperref}
\usepackage[labelformat=simple]{subcaption}
\renewcommand\thesubfigure{(\alph{subfigure})}
\usepackage{cancel}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{titling}

\setlength{\droptitle}{-13em}

@@

\title{Towards a Wide-Coverage Grammar: \\ Graphical Abstract Categorial Grammars}
\author{Jiří Maršík \and Maxime Amblard}
\maketitle

\begin{abstract}
  We present work whose ultimate goal is the creation of a wide-coverage
  abstract categorial grammar (ACG) that could be used to automatically
  build discourse-level representations. We focus on the challenge of
  integrating the treatment of disparate linguistic constraints in a
  single ACG and propose a generalization of the formalism: graphical
  abstract categorial grammars.
\end{abstract}

\section{Motivation}

Abstract categorial grammars (ACGs) have shown to be a viable formalism
for elegantly encoding the dynamic nature of discourse. Proposals based
on continuation semantics \cite{de2006towards} have tackled topics such
as event anaphora \cite{qian2011event}, SDRT discourse structure
\cite{asher2011sdrt} and modal accessibility constraints
\cite{asher2011montagovian}. However, all of these treatments only
consider tiny fragments of languages. We are interested in building a
wide-coverage grammar which integrates and reconciles the existing
formal treatments of discourse and allows us to study their interactions
and to build discourse representations automatically.

In this paper we focus on the problem of constructing a wide-coverage
ACG. We identify a problem that comes up when one tries to enforce
multiple constraints/dependencies in a single ACG and we propose a
generalization of the formalism which allows for a natural solution to
the problem.

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/double-acg.pdf}
    \caption{\label{fig:acg-comp-basic} Connecting form with meaning.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/serial-over-parallel.pdf}
    \caption{\label{fig:acg-comp-constr} Adding a constraint.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/parallel-over-serial.pdf}
    \caption{\label{fig:acg-comp-sem} Distinguishing syntactic and
      semantic ambiguities..}
  \end{subfigure}
  \caption{\label{fig:acg-comp} Diagrams of systems of ACGs.}
\end{figure*}

\section{The Problem of Multiple Constraints}
\label{sec:constraints}

We will consider several linguistic constraints that have been given
formal treatments in grammatical formalisms.

In French, negation is signalled by prepending the particle \emph{ne} to
the negated verb in conjunction with using a properly placed
accompanying word, such as a negative determiner in one of the verb's
arguments. This phenomenon has been elegantly formalized in the Frigram
interaction grammar \cite{perrier2007french}.

\begin{exe}
  \ex \label{ex:aucun-shallow} Jean \textbf{ne} parle à \textbf{aucun} collègue. \\
      (Jean speaks to no colleague.)
  \ex \label{ex:aucun-deep-obj} Jean \textbf{ne} parle à la femme d'\textbf{aucun} collègue. \\
      (Jean speaks to the wife of no colleague.)
  \ex \label{ex:aucun-deep-subj} \textbf{Aucun} collègue de Jean \textbf{ne} parle à sa femme. \\
      (No colleague of Jean's speaks to his wife.)
\end{exe}

We see here that the negative determiner \emph{aucun} can be present in
the subject or the object of the negated verb and it can modify the
argument directly or be attached to one of its complements. Furthermore,
we note that omitting either the word \emph{ne} or the word \emph{aucun}
while keeping the other produces a sentence which is considered
ungrammatical.

This difference in syntactic behavior between noun phrases that contain
a negative determiner and those that do not has implications for our
grammar. Since two terms having an identical type in an ACG signature
can be freely interchanged in any sentence, we are forced to assign two
different types to these two different kinds of noun phrases.

% NOTE: We could use a single constant for the two paired words.
% NOTE: We could use a higher-order type for the negative determiner
% which demands the negative particle itself.

This leads us to a grammar in which we subdivide the atomic types $N$
and $NP$ into subtypes that reflect whether or not they contain a
negative determiner inside. Types of the other constants, such as the
preposition \emph{de} seen in (\ref{ex:aucun-deep-obj}) and
(\ref{ex:aucun-deep-subj}), will constrain their arguments to have
compatible features on their types and will propagate the information
carried in the features to its result type, e.g.:

\begin{align*}
N_{de_1} &: NP\_NEG{=}F \limp N\_NEG{=}F \limp N\_NEG{=}F \\
N_{de_2} &: NP\_NEG{=}F \limp N\_NEG{=}T \limp N\_NEG{=}T \\
N_{de_3} &: NP\_NEG{=}T \limp N\_NEG{=}F \limp N\_NEG{=}T
\end{align*}

Enforcing other constraints leads us to subdivide our ``atomic'' types
even further (e.g. the authors of \cite{pogodalla2012controlling} add
features to the $S$ and $NP$ types to implement constraints about
extraction). Other phenomena, such as agreement on morphological
properties like number, gender, person or case, intuitively lead us to
make our types even more specific.

If we were to use this approach to write a grammar that enforces
multiple constraints at the same time, we would end up with complicated
types like the one below which provide complete specifications of the
various possible situations.

\begin{align*}
C_{de_{11}} :\ &(NP\_\textcolor{red}{NEG{=}T}\_\textcolor{green}{VAR{=}F}\_\textcolor{blue}{NUM{=}PL}) \\
&\limp (N\_\textcolor{red}{NEG{=}F}\_\textcolor{blue}{NUM{=}SG}) \\
&\limp (N\_\textcolor{red}{NEG{=}T}\_\textcolor{blue}{NUM{=}SG})
\end{align*}

This creates two problems. Firstly, the number of such types grows
exponentially with the number of features added. This can be fixed by
introducing dependent types into the type system as in
\cite{de2007type}. Secondly, and perhaps more importantly, we have lost
the clarity and simplicity of describing the individual phenomena
independently. This makes adding new constraints more and more
cumbersome and reduces our capability to reason about our grammar. It no
longer becomes apparent which language mechanisms behave independently
and their interactions are hard to track. We firmly believe that
simplicity is a fundamental requirement for constructing a large and
robust grammar and our proposal aims to reclaim that simplicity.

\section{Systems of ACGs}
\label{sec:acg-comp}

An ACG is defined in terms of two higher-order signatures, sets of typed
constants on which we can form well-typed lambda terms, and a
translating function from terms of one of the signatures to terms of the
other (the former signature being called the \emph{abstract} signature,
the latter being called the \emph{object} signature and the function
being called a \emph{lexicon}). We are then interested in the terms of
the abstract signature having some distinguished type and their images
through the lexicon (these two sets of terms are called the
\emph{abstract language} and the \emph{object language}, respectively).

When describing language, we are usually interested in systems of more
than one ACG and we use diagrams such as the ones on
Figure~\ref{fig:acg-comp} to explain them.

In Figure~\ref{fig:acg-comp-basic}, we have abstract syntactic terms
from the signature $\Sigma_{Synt}$ yielding the phonological strings and
semantic representations using two different lexicons. In
Figure~\ref{fig:acg-comp-constr}, we extend this pattern by adding a new
abstract signature which narrows the language of syntactic terms and
implements some linguistic constraint. Finally, in
Figure~\ref{fig:acg-comp-sem}, we split the abstract signature
$\Sigma_{Synt}$ of Figure~\ref{fig:acg-comp-basic} into two signatures,
$\Sigma_{Syntax}$ and $\Sigma_{SyntSem}$, in order to model syntactic
and semantic ambiguities in different grammars.

In our grammar, we would like to combine both of the patterns on
Figures~\ref{fig:acg-comp-constr} and \ref{fig:acg-comp-sem} and
furthermore, to use the pattern of Figure~\ref{fig:acg-comp-constr}
multiple times to implement multiple constraints independently. However,
these combinations are not possible in the framework of ACGs without
introducing complexity such as by contaminating the syntax-semantics
interface by the implementation details of the syntactic layer.

We would like to have a system which would be characterized by a diagram
like the one on Figure~\ref{fig:gacg}. However, ACG diagrams are limited
to arborescences and we are obliged to generalize the formalism in order
to get the expected interpretation of Figure~\ref{fig:gacg}.

\section{Graphical ACGs}

\begin{figure}
  \centering
  \includegraphics[width=0.25\textwidth,height=0.2\textheight]{../diagrams/final.pdf}
  \caption{{\label{fig:gacg} A graphical ACG that implements two
      independent syntactic constraints and distinguishes syntactic and
      semantic ambiguities.}}
\end{figure}

We define a \emph{graphical abstract categorial grammar} as a directed
acyclic graph (DAG) whose nodes are labeled with signatures (and
distinguished types) and whose edges are labeled with lexicons, a
mathematical reification of an ACG diagram generalized from
arborescences to DAGs. We then search for an appropriate semantics for
these structures, i.e. how to determine what languages are defined by a
graphical ACG.

We first follow the paradigm in which we associate languages with the
nodes of the graph while the edges tell us how these languages are
defined in terms of each other. We argue that intersection of languages
is the suitable meaning of more than two edges arriving at the same node
based both on the simplicity of the resulting definitions and on our
expectations about the desired semantics. This leads us to the following
definitions of \emph{intrinsic} and \emph{extrinsic} languages
associated with some node $v$ in a graphical ACG $\mathcal{G}$:

$$
\mathcal{I}_{\mathcal{G}}(v) = \{t \in \Lambda(\Sigma_v)
\mid\ \vdash_{\Sigma_v} t : S_v\}
$$
$$
\mathcal{E}_{\mathcal{G}}(v) = \mathcal{I}_{\mathcal{G}}(v) \cap
\bigcap_{(u,v) \in E} \mathcal{L}_{(u,v)}(\mathcal{E}_{\mathcal{G}}(u))
$$

The intrinsic language is just the set of terms built on the node's
signature and having the node's distinguished type. The extrinsic
language is established by taking the extrinsic languages of its
predecessors, mapping them through a lexicon and taking their
intersection, or just taking the node's intrinsic language if it has no
predecessors.

We examine the relationship between the languages defined by ACGs and
graphical ACGs (G-ACGs). Intrinsic languages correspond exactly to
abstract languages and therefore the sets of languages definable by both
are equal.

$$
\mathcal{I} = \mathcal{A}
$$

G-ACG extrinsic languages correspond to ACG object languages with
intersection. More formally, we can say that ACG object languages are
just ACG abstract languages closed on transformation by a lexicon and we
show that G-ACG extrinsic languages are just ACG abstract languages
closed on transformation by a lexicon and intersection.

$$
\mathcal{O} = \mathcal{A}^{\mathcal{L}}
$$
$$
\mathcal{E} = \mathcal{A}^{\mathcal{L}{\cap}}
$$

We then argue that closure on intersection is an intuitive property of a
grammatical formalism that allows independent combination of linguistic
constraints. Furthermore, from the above we can see that object
languages are as expressive as extrinsic languages iff object languages
are closed on intersection, which is conjectured to be false.

However, extrinsic languages turn out to give counter-intuitive
interpretations to some graphical ACGs and so after examining the
algebraic paradigm which gave us the intrinsic and extrinsic languages,
we consider an alternative paradigm, which we call generative.

In this paradigm, we identify the nodes of the graph with terms and the
edges with statements that one term is mapped into another using a
lexicon. This leads us to the definition of the \emph{pangraphical}
language of a node $u$ in a G-ACG $\mathcal{G}$.

A term $t$ belongs to $\mathcal{P}_{\mathcal{G}}(u)$ whenever there
exists a labeling $T$ of the nodes of the graph such that:

\begin{itemize}
  \item $T_u = t$.
  \item For all $v \in V(G)$, $\vdash_{\Sigma_v} T_v : S_v$.
  \item For all $(v,w) \in E(G)$, $\mathcal{L}_{(v,w)}(T_v) = T_w$.
\end{itemize}

The two paradigms under which we can interpret ACGs are equivalent in
the context of plain ACGs and arborescent diagrams and the two metaphors
(nodes as languages and nodes as terms) can be used interchangeably. We
back this up formally by showing that for every node $u$ in every
arborescent G-ACG $\mathcal{G}$, we have $\mathcal{E}_{\mathcal{G}}(u) =
\mathcal{P}_{\mathcal{G}}(u)$. However, as we start to consider
non-arborescent graphs, we find, interestingly, that the two paradigms
diverge.

The newly defined pangraphical languages solve the problem of extrinsic
languages giving us counter-intuitive interpretations for some specific
G-ACGs. We show by construction that pangraphical languages are as
expressive as extrinsic languages and so the kinds of languages we have
defined here are increasingly more expressive, $\mathcal{I} \subseteq
\mathcal{E} \subseteq \mathcal{P}$. We also show that
$\mathcal{I}_{\mathcal{G}}(u) \supseteq \mathcal{E}_{\mathcal{G}}(u)
\supseteq \mathcal{P}_{\mathcal{G}}(u)$, meaning that the language
definitions are more and more constraining and that the extrinsic
languages went in the right direction as pangraphical languages, but not
as far.

Finally, we construct a G-ACG which integrates the French negation
constraint discussed in Section~\ref{sec:constraints}, the constraints
on extraction introduced in \cite{pogodalla2012controlling} and a
constraint handling gender agreement in a single grammar
specification. The negation constraint is implemented exactly as was
described in Section~\ref{sec:constraints} and the extraction
constraints are taken as is from \cite{pogodalla2012controlling}. The
underlying syntactic signature is defined without any concern for the
additional constraints and the syntax-semantics signature is kept simple
as well.

In the end, this lets us define the syntax in a clean way using the
idiomatic style of categorial grammars (simple atomic types like $N$,
$NP$ and $S$) and then define the constraints themselves the same way
they are defined in papers which try to formalize them individually
(such as the case with \cite{pogodalla2012controlling}). We can freely
combine both the constraint pattern and the syntax-semantics ambiguities
pattern from Section~\ref{sec:acg-comp}.

\bibliographystyle{plain}
\bibliography{../biblio}
